---
title: "Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Integration) 
library(tidyverse)
```


# Overview 

The Integrate package implements integration using the Newton-Cotes algorithm.
It focuses on integrating using the following rules:

- Riemann integration
  - Riemann integration is when we approximate the area under a curve using a 
  sequence of rectangles whose heights are computed using points from the left hand side, 
  right hand side, or middle of the interval that composes a rectangle
  - It is defined as the limit to the approximated sum
  - $\int_a^b f(x)~dx = \lim_{n\to \infty} \sum_{i=1}^n \frac{b-a}{n} f(x_i)$
- Trapezoidal rule
  - This is a more sophisticated way to approximate the shape of the function that 
  is within the subinterval. Instead of using rectangles, we use lines instead (ie linear interpolation)
  - The following equation is created:
  $\int_a^b f(x)~dx \approx \sum_{i=0}^n \frac{f(a+i/n) + f(a+(i+1)/n)}{2} \cdot \frac{b-a}{n}$
- Simpson's rule
  - An even more complex way to approximate the area under a curve, and it is done
  using quadratic interpolation
  - In this case we find the value of the quadratic function in each subinterval
  that will connect the end points **and** also interpolate the middle
  - The resulting equation is:
  $\int_a ^b f(x) ~dx \approx \sum_{i=1}^n \frac{x_{i}-x_{i-1}}{6}\left[ f(x_{i-1}) + 4f\left(\frac{x_{i} + x_{i-1}}{2}\right) + f(x_{i}) \right].$


# Usage

In order to use the functions in the Integrate package, a few key parameters are
required.

`f` = function of your choosing that has either **one** input value, or one input
value and the other arguments have default values (ie dexp function). It must also 
only return one value.

`range` = a vector of length 2 that species the range you want to integrate over. 
So range = c(0,10) means you want to find the area under the curve from 0 to 10

`n` = the number of intervals you would like to use to make the approximation. Note
that the larger n is, the better the approximation to the area is!

## Unique Function

Here is an example of how to use the Integration package on an in-built function.
Simpson's method is the most accurate.
```{r}
resultRiemann <- Integration::riemann(100,sqrt,c(0,10))
resultTrapezoid <- Integration::trapezoid(100,sqrt,c(0,10))
resultSimpsons <- Integration::simpsons(100,sqrt,c(0,10))

paste("The approximate Riemann value is", resultRiemann)
paste("The approximate Trapezoidal value is", resultTrapezoid)
paste("The approximate Simpsons value is", resultSimpsons)
```

Another example, if one uses a function built on their own. Once again, we see
that simpsons method is the most accurate

```{r}
cubicFunction = function(x){
  result = x^3
}
```

```{r}
resultRiemann <- Integration::riemann(100,cubicFunction,c(0,2))
resultTrapezoid <- Integration::trapezoid(100,cubicFunction,c(0,2))
resultSimpsons <- Integration::simpsons(100,cubicFunction,c(0,2))

paste("The approximate Riemann value is", resultRiemann)
paste("The approximate Trapezoidal value is", resultTrapezoid)
paste("The approximate Simpsons value is", resultSimpsons)
```

## Comparison 

Below you will find comparisons of the various rules and see how their efficacy 
holds up compared to each other.

### Graphs

```{r}
# load functionvalues
x = 0:20
f <- sqrt(x)
f <- data.frame(f = f, 
                   x = x)

# set params
a = 0
b = 10
nIntervals = 5
```

#### Target Area

The graph below illustrates the target area under the curve.
```{r}
df1 <- tibble(x = unique(sort(c(seq(0,b,.01),20))),
                    f = sqrt(x))

df2 <- df1 %>%
  filter(x <= b & x>=a) %>%
  bind_rows(tibble(x = b, f = 0)) %>%
  bind_rows(tibble(x = a, f = 0)) 

finalFig <- df1 %>%
  ggplot(aes(x = x, y = f)) + 
  geom_line(colour = "red") +
  ylab("Density") + 
  geom_hline(yintercept = 0)+
  geom_polygon(data = df2, fill = "red", alpha = .5)+
  theme_minimal()+ 
  labs(title = "Target Area of Square Root Function", 
       subtitle = "Integrate from 0 to 10")

finalFig

```



#### Riemann

```{r}
dfRiemann = tibble(n = nIntervals, 
                   x = seq(a,b,length = nIntervals+1)[-(nIntervals+1)], 
                   f = sqrt(x))%>%
  mutate(x1 = x + (b-a)/(2*n))

finalFig+
  geom_bar(data = dfRiemann, aes(x=x1,y=f),stat = "identity",fill="darkblue",
           alpha=0.7,
           width = (b-a)/nIntervals)+
  theme_minimal()+ 
  labs(title = "Riemann Integration of Square Root Function", 
       subtitle = "Riemann Approximations in Blue")
```
```{r}
resultRiemann <- Integration::riemann(nIntervals,sqrt,c(a,b))
paste("The approximate Riemann value with range (",a,",",b,")", "and",nIntervals,"intervals is", round(resultRiemann,3))
```

#### Trapezoidal Rule
We can see that compared to Riemann integration, the trapezoid rule covers a lot
more of the target area. 
```{r}
dfTrapezoid = tibble(n = nIntervals, 
                   x = seq(a,b,length = nIntervals+1), 
                   f = sqrt(x))%>%
  bind_rows(tibble(n = nIntervals, 
                   x = b-a,
                   f = 0))

finalFig+
  geom_polygon(data= dfTrapezoid, aes(x=x,y=f),fill="darkblue",alpha=0.7)+
  theme_minimal()+ 
  labs(title = "Trapezoidal Integration of Square Root Function", 
       subtitle = "Trapezoidal Approximations in Blue")
```

```{r}
resultTrapezoid <- Integration::trapezoid(nIntervals,sqrt,c(a,b))
paste("The approximate Trapezoidal value with range (",a,",",b,")", "and",nIntervals,"intervals is", round(resultTrapezoid,3))
```

It will be difficult to show how the area is integrated using Simpson's rule 
as it is more complex. However the speed of integration between the Riemann method, 
Trapezoid Rule and Simpson's rule will be compared in the later section.

### Speed

Here we compare the efficacy of the various algorithms in computing the integral 
of the square root function. 

We can see that when Simpson's rule is used, it takes the least amount of time
to converge to the correct answer, followed by the trapezoid rule. Using Riemann 
integration is not as efficient since it takes at least 50 iterations for it to be 
close to the correct area, and it consistently underestimates the area under the 
curve. 
```{r}
n = 1:100
dfSpeed = tibble(n =n)
dfSpeed$Riemann = sapply(dfSpeed$n, Integration::riemann, sqrt, c(a,b))
dfSpeed$Trapezoid = sapply(dfSpeed$n, Integration::trapezoid, sqrt, c(a,b))
dfSpeed$Simpsons = sapply(dfSpeed$n, Integration::simpsons, sqrt, c(a,b))
```

```{r}
dfSpeedLong = dfSpeed %>%
  pivot_longer(cols = c("Riemann","Trapezoid","Simpsons"), names_to = "Algorithm", 
               values_to = "Values")

ggplot(dfSpeedLong)+ 
  geom_line(aes(x=n,y = Values, color = Algorithm))+ 
  theme_minimal()+ 
  labs(title = "Comparing Covergence of Integration Algorithms",
       subtitle = "Integrating the square root function from 0 to 10",
       x = "Number of Intervals",
       y = "Approximated Area Under the Curve")
```




